Method to infer relatedness using low coverage ancient DNA

### 2022-02-01 paper_overleaf/main.tex
Updates from Overleaf

We see that in this case, the siblings model predictions match true IBD states the most. This is reflected in the log likelihoods values of the models as well, where siblings model has the highest value. We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48). All the models in fig.1 utilize information about ROH given by the ROH-HMM.

@@

We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in fig.3. Here, and in fig.4, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'. Unrelated label here refers to KIn performance results when all unrelated, degree5, degree4, degree3 pairs are labelled as unrelated. This is useful for comparison to READ, since READ classifies relatedness upto degree2. 'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals. Fig.3 shows that for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used. For second degree, at 0.1x we see a high false positive rate, that does not go down much with an increase in cutoff. For the same relationship, at a coverage of 0.2x, a cutoff of 1 seems appropriate to reduce false positive rate to below $5\%$. Similarly, for siblings at 0.1x, a cutoff of 1 works well to keep false positives low. For parent-child, we find that a cutoff of about 2 is a better choice. Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of around $5\%$. In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between paret-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.

@@

We show that READ outputs higher false positives in case of recent inbreeding in the population, and higher false negatives in case of contamination.

@@

-supplementary table2
\begin{figure}[h]
    \centering
    \includegraphics[width=18cm]{supplementary_info/plots/}
    \caption{Avuncular and Grandparent-grandchild form different clusters, when total number of IBD fragments are plotted against total length of IBD fragments.}
    \label{figS5}
\end{figure}

@@

Methods
$$\sum_{w} \sum_{I}\sum_\omega \log P(D_{w}|Z_w, H_w)k_{w\omega i}$$

>>>>

We see that in this case, the siblings model predictions match true IBD states the most. This is reflected in the log likelihoods values (LL) of the models as well, where siblings model has the highest value. We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48). All the models in fig.1 utilize information about ROH given by the ROH-HMM.

@@

We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in fig.3. Here, and in fig.4, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'. Unrelated label here refers to KIn performance results when all unrelated, degree5, degree4, degree3 pairs are labelled as unrelated. This is useful for comparison to READ, since READ classifies relatedness upto degree2. 'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals. Fig.3 shows that for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used. For second degree, at 0.1x we see a high false positive rate, that does not go down much with an increase in cutoff. For the same relationship, at a coverage of 0.2x, a cutoff of 1 seems appropriate to reduce false positive rate to below $5\%$. Similarly, for siblings at 0.1x, a cutoff of 1 works well to keep false positives low. For parent-child, we find that a cutoff of about 2 is a better choice. Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of around $5\%$. In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between parent-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.

@@

Several methods are available to estimate relatedness from modern as well as ancient DNA data. We develop a new method to estimate genetic kinship from low coverage data. We show that it is possible to accurately infer relatedness, along with IBD tracts, from such data in presence of long ROH, contamination, as well as ascertainment bias. Our method utilizes Hidden Markov Models to estimate IBD tracts, conditional on a particular relatedness, and we use log likelihoods of these HMMs to find the best fit. This approach of using Hidden Markov Model makes use of the fact that DNA is inherited in blocks, and hence the IBD state in a given window is dependent on the IBD state in the previous window. We use this approach to identify the following relatives: Identical individuals, Parent-Child, Siblings, Second degree, Third degree, Unrelated individuals. We show that it is possible to differentiate between avuncular and grand-parent-grandchild relations. We have developed a similar Hidden Markov Model to infer long ROH, and use the posterior probabilities to make better relatedness inference. We further use a simple model to correct the pairwise differences in genomic windows for contamination.

We demonstrate using simulations, that our method performs well at low coverage. We compared KIn's model performance to a widely used software READ to show that our approach works at least as good as READ on ideal data without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives. We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate. Contrary to the intuition, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIn and READ. READ's performance improves because ROH does not cause any change in pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals. More is the relatedness, the more is the decrease in pairwise differences in presence of ROH. And so long ROH causes higher power for READ which otherwise is a conservative method. KIn performs well for the same reasons, but can miss-classify second degree individuals with ROH as siblings, if ROH is not accounted for. Hence, we implement ROH states in the relatedness model to avoid such miss classifications. Contamination, on the other hand can make the performance of READ much worse depending on the amount of contamination, and the number of individuals contaminated. We show that even at contamination levels  $<=3\%$, and contamination in only 8 out of 17 individuals, READ performance falls dramatically. Contamination in more individuals will cause a shift in median difference in the population and may case non contaminated individuals to look more related, thus increasing the false positive rates. On the other hand, increase in amount of contamination in few individuals can make them look more unrelated. KIn, with the implementation of a model to account for contamination performs well in this case. Ascertainment causes a decrease in power for both the methods, since it reduces the amount of data. However, the decrease in power is more for READ compared to KIn.

DNA data from Bronze age, Lech valley has low contamination, and no ROH. We demonstrate that KIn matches the results from READ and lcMLkin for pairs with good amount of overlaps, while adds more information for low coverage samples. As expected from the simulations, KIn is helpful in differentiating between parent-child from siblings, and identifying second and third degree individuals in cases where there is an overlap of only few thousand sites or less, and the other methods are not informative. We show that when applied to neandertal specimens from Chagyrskaya cave, READ and KIn agree in relatedness estimates for all pairs where both methods have sufficient data. KIn additionally identifys a pair of first degree relatives as parent-child, which is also confirmed with mtDNA. In addition, KIn identifies a third degree relative. We compare IBD estimates from KIn to lcMLkin and show that KIn's IBD estimates are much closer to the expected estimates, given the relatedness between individuals that has been estimated with both READ and KIn. 
Output of KIn is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio. Due to the tabular output, KIn is easy to automatise and apply on large datasets. 
One limitation with our approach is that this method assumes a single population. In case of population structure, with the subpopulations with very different diversity, KIn may show inaccurate inferences. Our method makes another assumption, that the dataset comprises of at least some unrelated individuals. We estimate $p_0$ as the median of all pairwise differences, and if more than half of the pairs are related to each other, $p_0$ estimate would be biased, and may affect the relatedness inference. In this study, we identify six different cases of relatedness, but in theory our approach can be extended to identify many more such cases, such as double first cousins, using a corresponding IBD state transition matrix.

@@

-supplementary table2
\begin{figure}[h]
    \centering
    \includegraphics[width=18cm]{supplementary_info/plots/egplot1.png}
    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIn differs from READ. (A) AITI62B-OTTM156 (READ estimates Identical individuals, lcMLkin does not have sufficient data, and KIn predicts parent-child. (B) UNTA5867-UNTA5868Sk1 (READ estimates Second Degree, lcMLkin does not have sufficient data, and KIn predicts parent-child). (C) AITI40-AITI72 (READ estimates Unrelated individuals, lcMLkin shows 3rd-5th degree, and KIn predicts 3rd degree.)
    (D) AITI2-AITI55 (READ estimates Unrelated individuals, lcMLkin and KIn predicts predict second degree.)}
    \label{figS5}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=18cm]{supplementary_info/plots/egplot2.png}
    \caption{Plots showing proportion of differences in windows along the genome for some of the pairs of relatives for which KIn differs from lcMLkin. (A) AITI43-AITI55 (READ estimates First degree, lcMLkin predicts siblings, while KIn predicts parent-child. (B) AITI70-AITI72 (READ estimates First degree, lcMLkin predicts parent-child, while KIn predicts siblings.}
    \label{figS5}
\end{figure}

@@

Discussion

Our contamination correction model requires an estimate of divergence between the contaminating and the target populations. We provide the users an option to provide a file with average pairwise difference between the two populations, or to provide a vcf file with an individual from each population. 
Pipeline has p(diff) instead of pseudohap
Methods
$$\sum_{w} \sum_{I}\sum_\omega \log P(D_{w}|Z_w, H_w)k_{w\omega i}$$
